\documentclass[11pt, a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{top=1in, bottom=1in, left=1in, right=1in}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{mdframed}
\usepackage{minted}
\usepackage{ifthen}

\newdimen\longline
\longline=\textwidth\advance\longline-6cm
\def\LayoutTextField#1#2{#2}
\def\lbl#1{\hbox to 5cm{#1\hfill\strut}}%
\def\labelline#1#2#3{\lbl{#1}\vbox{\hbox{\TextField[bordercolor=white,name=#2,width=#3]{\null}}\kern2pt\hrule}}
\def\q#1#2{\hbox to \hsize{\labelline{#1}{#2}{\longline}}\vskip1.4ex}



% --- Code Listing Styling ---
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.96}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{blue},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=none,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=4
}
\lstset{style=mystyle}

\title{\textbf{Lab Exercise 1: Measuring the Syscall Tax}}
\author{Gianmaria Del Monte}
\date{}

\begin{document}

\maketitle

\section*{Background}
During the lecture, we discussed the hidden cost of traditional I/O. Every time your application invokes a standard \texttt{read()} or \texttt{write()} command, it forces a context switch (a mode switch from user space to kernel space). While a single mode switch is relatively fast (a fraction of a microsecond), doing it millions of times will severely degrade your application's throughput. 

In this exercise, you will use a provided C program and the Linux \texttt{strace} utility to visually quantify the ``syscall tax.'' You will observe how reading a large file with drastically different buffer sizes impacts the number of system calls and the overall execution time.

\section*{Part 1: Environment Setup}
First, you need to download the source code for this lab from the course repository. Open your terminal and run the following commands to clone the repository and navigate to the correct directory:

\begin{minted}{shell}
    git clone https://github.com/gmgigi96/icscexercises.git
    cd icscexercises/exercise1
\end{minted}

Next, let's create a large dummy data file to read from. We will use the \texttt{dd} utility to generate a 500 MB file filled with random data (this might take a few seconds to complete):

\begin{minted}{shell}
    dd if=/dev/urandom of=large_file.bin bs=1M count=500
\end{minted}

Finally, compile the C program using the provided Makefile:

\begin{minted}{shell}
    make
\end{minted}

\section*{Part 2: The Experiment}
We will now test how buffer size changes the OS behavior using \texttt{strace}, a diagnostic tool that intercepts and records system calls. We use the \texttt{-c} flag to get a statistical summary.

\subsection*{Task 2.1: The 1-Byte Buffer (The Worst-Case Scenario)}
Run the program reading just 1 byte at a time. This forces the CPU to switch from user space to kernel space for every single byte in the 500 MB file.
\begin{minted}{shell}
    strace -c ./read_test large_file.bin 1
\end{minted}

\textit{Observe the output table. How many \texttt{read} calls were executed?}


\subsection*{Task 2.2: The 4KB Buffer (The Block Size)}
Now, let's match the typical logical block size of a storage drive and a standard RAM page (4096 bytes).
\begin{minted}{shell}
    strace -c ./read_test large_file.bin 4096
\end{minted}

\textit{Observe how drastically the execution time dropped and the new number of \texttt{read} syscalls.}

\subsection*{Task 2.3: The 1MB Buffer (The Batching Approach)}
Finally, let's give the kernel a massive buffer to work with.
\begin{minted}{shell}
    strace -c ./read_test large_file.bin 1048576
\end{minted}

\textit{Observe the throughput. Is the performance gap between 4KB and 1MB as massive as the gap between 1 Byte and 4KB?}

\subsection*{Task 2.4: The C Standard Library Rescue (User-Space Buffering)}

In Task 2.1, you saw that reading a file 1 byte at a time using the raw POSIX \texttt{read()} system call destroyed performance due to the syscall tax. However, sometimes application logic \textit{requires} processing a file character-by-character (e.g., parsing a CSV). 

To solve this without forcing the programmer to write complex batching logic, the C Standard Library (\texttt{libc}) provides user-space buffering via functions like \texttt{fopen()} and \texttt{fread()}. 

We have provided a second program, \texttt{read\_libc.c}. Look at its source code. It reads the file exactly 1 byte at a time, just like Task 2.1, but it uses \texttt{fread()} instead of \texttt{read()}. Let's trace it:

\begin{minted}{shell}
    strace -c ./read_libc large_file.bin
\end{minted}
\textit{Observe the output. Even though the C code loops 500 million times requesting 1 byte, how many actual \texttt{read} system calls did the OS kernel execute? What does this tell you about what \texttt{libc} is doing behind the scenes in user space?}

\section*{Part 3: Lab Questions}

\noindent \textbf{Question 1:} In the 1-byte buffer test, your SSD/HDD was likely barely active, yet the program took a long time. If the physical disk wasn't the bottleneck, what exact component or process was?

\vspace{0.2cm}
\noindent
\fbox{
    \begin{minipage}{0.97\textwidth}
        \vspace{4.5cm} % Adjust this to make the box taller or shorter
        \hfill
    \end{minipage}
}

\vspace{0.8cm}

\noindent \textbf{Question 2:} When you increased the buffer from 4KB to 1MB, the number of syscalls decreased, but the overall throughput might not have scaled perfectly linearly. Based on the ``Page Cache'' concept from the lecture, why do you think diminishing returns kick in at very large user-space buffer sizes?

\vspace{0.2cm}
\noindent
\fbox{
    \begin{minipage}{0.97\textwidth}
        \vspace{5.5cm} % Adjust this to make the box taller or shorter
        \hfill
    \end{minipage}
}

\end{document}